Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-26 paper.qmd
Updated Figure 3

With our goal being to make a representation of this \gD{} layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first bin the data, reducing it from $n$ to $m\leq n$ observations, and connect the bin centroids. We recommend using a hexagon grid because it better reflects the data distribution and has less artifacts than a rectangular grid. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-scurve, and documented below.

@@

umap_two_curvy_scaled <- two_curvy_scaled_obj$scaled_nldr
lim1 <- two_curvy_scaled_obj$lim1
lim2 <- two_curvy_scaled_obj$lim2
r2 <- diff(lim2)/diff(lim1)

sc_ltr_pos <- c(0.08, 0.93)
# sc_xlims <- c(-0.5, 1.47)
# sc_ylims <- c(-0.32, 2.1)
sc_xlims <- c(-0.25, 1.25)
sc_ylims <- c(-0.3, 2)

nldr_scurve <- umap_scurve_scaled |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos)
```

@@

```{r}
#| label: hexbin-scurve
## Compute hexbin parameters
num_bins_x_scurve <- 7

## hexagon binning to have regular hexagons
hb_obj_scurve <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = num_bins_x_scurve, 
  r2 = r2,
  q = 0.1)

## Data set with all centroids
all_centroids_df <- hb_obj_scurve$centroids

## Generate all coordinates of hexagons
hex_grid <- hb_obj_scurve$hex_poly

## To obtain the standardise counts within hexbins
counts_df <- hb_obj_scurve$std_cts
df_bin_centroids <- extract_hexbin_centroids(
  centroids_df = all_centroids_df, 
  counts_df = counts_df) |>
  filter(drop_empty == FALSE)

@@

hex_grid_scurve <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "black", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5, size = 0.5, color = "#000000") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos)
```

@@

```{r}
#| label: empty-bin-scurve
hex_grid_nonempty <- hex_grid |>
  filter(hex_poly_id %in% df_bin_centroids$hexID)

hex_grid_nonempty_scurve <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(color = "black", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids, 
             aes(x = c_x, y = c_y), 
             color = "#33a02c",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos) 
```

<!--2D model-->

```{r}
#| label: triangulate-scurve
## Triangulate bin centroids
tr1_object_scurve <- tri_bin_centroids(
  df_bin_centroids, x = "c_x", y = "c_y")
tr_from_to_df_scurve <- gen_edges(
  tri_object = tr1_object_scurve)

## Compute 2D distances
distance_scurve <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_scurve, 
  start_x = "x_from", 
  start_y = "y_from", 
  end_x = "x_to", 
  end_y = "y_to", 
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_scurve <- find_lg_benchmark(
  distance_edges = distance_scurve, 
  distance_col = "distance")

trimesh_removed_scurve <- vis_rmlg_mesh(
  distance_edges = distance_scurve, 
  benchmark_value = benchmark_scurve, 
  tr_coord_df = tr_from_to_df_scurve, 
  distance_col = "distance") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("d", sc_ltr_pos) 
```

@@

```{r}
#| label: s-curve-true-model-proj1

true_model_df <- read_rds("data/s_curve/scurve_true_model.rds")
wireframe_true_model <- read_rds("data/s_curve/scurve_true_model_wireframe.rds")

data_scurve <- training_data_scurve |> 
  select(-ID) |> 
  mutate(type = "data")

model_scurve <- true_model_df |> 
    select(-ID) |> 
  mutate(type = "model")

# Apply the scaling
df_model_data_scurve <- bind_rows(data_scurve, model_scurve)
scaled_scurve <- scale_data_manual(df_model_data_scurve, "type") |>
  as_tibble()

scaled_s_curve_noise <- scaled_scurve |>
  filter(type == "data") |>
  select(-type)

scaled_s_curve_true <- scaled_scurve |>
  filter(type == "model") |>
  select(-type)

projection <- cbind(
    c(0.45181,0.04747,0.02613,0.25735,0.20622,0.08755,-0.11488),
    c(0.03231,0.23556,0.35166,-0.11741,0.12730,0.09544,0.34262))

projected <- as.matrix(scaled_s_curve_noise) %*% projection

projected_df <- projected |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

projected_true <- as.matrix(scaled_s_curve_true) %*% projection

projected_true_df <- projected_true |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

model_df <- dplyr::left_join(
  wireframe_true_model, 
  projected_true_df, 
  by = c("from" = "ID"))

names(model_df)[3:NCOL(model_df)] <- paste0(names(projected_true_df)[-NCOL(projected_true_df)], "_from")

model_df <- dplyr::left_join(model_df, projected_true_df, by = c("to" = "ID"))
names(model_df)[(2 + NCOL(projected_true_df)):NCOL(model_df)] <- paste0(names(projected_true_df)[-NCOL(projected_true_df)], "_to")

limits <- 1
rng <- range(projected)
projected <- projected/max(abs(rng))
colnames(projected) <- c("P1", "P2")
projected <- data.frame(projected)
obs_labels <- as.character(1:nrow(scaled_s_curve_noise))

axis_scale <- limits/6
axis_pos <- -2/3 * limits

# axis_scale <- limits/4
# axis_pos <- -0.5 * limits

adj <- function(x) axis_pos + x * axis_scale
axes <- data.frame(x1 = adj(0), 
                   y1 = adj(0), 
                   x2 = adj(projection[, 1]), 
                   y2 = adj(projection[, 2]))

axis_labels <- colnames(scaled_s_curve_noise)
rownames(axes) <- axis_labels

# axis_scale <- 0.15 * limits
# axis_pos <- -0.6 * limits
# adj <- function(x) axis_pos + x * axis_scale
theta <- seq(0, 2 * pi, length = 50)
circle <- data.frame(c1 = adj(cos(theta)), 
                     c2 = adj(sin(theta)))

scurve_proj1_true_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#3182bd") + 
  geom_point(
        size = 0.8,
        alpha = 0.5,
        color = "#000000") +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed()

```{r}
#| label: s-curve-true-model-proj2

projection <- cbind(
    c(0.03627,0.15991,0.39768,-0.21032,-0.14886,-0.17589,0.23305),
    c(-0.34731,-0.08183,0.15819,0.31926,0.08105,0.10083,0.25627))

projected <- as.matrix(scaled_s_curve_noise) %*% projection

projected_df <- projected |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

projected_true <- as.matrix(scaled_s_curve_true) %*% projection

projected_true_df <- projected_true |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

model_df <- dplyr::left_join(
  wireframe_true_model, 
  projected_true_df, 
  by = c("from" = "ID"))

names(model_df)[3:NCOL(model_df)] <- paste0(names(projected_true_df)[-NCOL(projected_true_df)], "_from")

model_df <- dplyr::left_join(model_df, projected_true_df, by = c("to" = "ID"))
names(model_df)[(2 + NCOL(projected_true_df)):NCOL(model_df)] <- paste0(names(projected_true_df)[-NCOL(projected_true_df)], "_to")

limits <- 1
rng <- range(projected)
projected <- projected/max(abs(rng))
colnames(projected) <- c("P1", "P2")
projected <- data.frame(projected)
obs_labels <- as.character(1:nrow(scaled_s_curve_noise))

# axis_scale <- 2 * limits/3
# axis_pos <- -0.6 * limits
axis_scale <- limits/6
axis_pos <- -2/3 * limits

adj <- function(x) axis_pos + x * axis_scale
axes <- data.frame(x1 = adj(0), 
                   y1 = adj(0), 
                   x2 = adj(projection[, 1]), 
                   y2 = adj(projection[, 2]))

axis_labels <- colnames(scaled_s_curve_noise)
rownames(axes) <- axis_labels

# axis_scale <- 0.15 * limits
# axis_pos <- -0.6 * limits
# adj <- function(x) axis_pos + x * axis_scale
theta <- seq(0, 2 * pi, length = 50)
circle <- data.frame(c1 = adj(cos(theta)), c2 = adj(sin(theta)))

scurve_proj2_true_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#3182bd") +
  geom_point(
        size = 0.8,
        alpha = 0.5,
        color = "#000000") +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed()

```{r}
#| label: fig-scurve-true-sc
#| fig-cap: "Two views of the true model (blue lines) in $2\\text{-}D$ projections from $7\\text{-}D$, for the S-curve data (black points). The data is spread along the S shape, and does not vary much from this curve. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/I5GL23vLiw0>)." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5

scurve_proj1_true_model + scurve_proj2_true_model + 
  plot_layout(ncol = 2)
```

<!--langevitour with true model for S-curve--> 
<!-- ::: {#fig-scurve-true-sc layout-ncol="3" fig-pos="H"} -->

<!-- ![](figures/scurve/sc_true_1.png){width="150" fig-align="center"} -->

<!-- ![](figures/scurve/sc_true_2.png){width="150" fig-align="center"} -->

<!-- ![](figures/scurve/sc_true_3.png){width="150" fig-align="center"} -->

@@

```{r}
#| label: fig-NLDR-scurve
#| echo: false
#| fig-cap: "Key steps for constructing the model on the UMAP layout ($k=2$): (a) data, (b) hexagon bins, (c) bin centroids, and (d) triangulated centroids. The S-curve data is shown."
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%

nldr_scurve + hex_grid_scurve + 
  hex_grid_nonempty_scurve + 
  trimesh_removed_scurve +
  plot_layout(ncol = 4)
```

@@

Because we are working with distances between points, starting with data having a standard scale, e.g. [0, 1], is recommended. The default should take the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$ into account. When $k=2$, as in hexagon binning, the default range is $[0, y_{i,\text{max}}], i=1,2$, where $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = \frac{r_2}{r_1}$ (@fig-NLDR-scurve). If the NLDR aspect ratio is ignored then set $y_ {2,\text{max}} = 1$. <!-- \times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height ($a_1$) and width ($a_2$) of a regular hexagon.) The scaling of data should take the size of the hexagons into account, but choice of number of bins should. -->

@@

      y = -0.35, 

@@

      yend = -0.35), #-0.35),

@@

    annotate("text", x=0.5, y=-0.45, 

@@

      x = -0.25, 

@@

      xend = -0.25, 

@@

    annotate("text", x=-0.35, y=1, 

@@

    annotate("text", x=-0.18, y=-0.25, 

@@

When $k=2$, the number of bins along the $x$ and $y$ axes, $b_1$ and $b_2$, is computed by considering the scaling factors ${r_1, r_2}$, along with the offset along the axes ($q_1$, $q_2$), and the height ($a_1$) and width ($a_2$) of a regular hexagon (@fig-NLDR-scurve (b)). ($b_1 = \frac{r_1 + q_1}{a_2}$, and $b_2 = \frac{r_2 + q_2}{0.75 \times a_1}$) where $0.75 \times a_1$ accommodate to have hexagons fill the entire 2D space without leaving any gaps between them.)-->

@@

By default, the bin centroid is used for describing a hexagon (as done in @fig-NLDR-scurve (c)), but any measure of center, such as a mean or weighted mean of the points within each hexagon, could be used. The bin centers, and the binned data, are the two important components needed to render the model representation in high dimensions.

@@

Delaunay triangulation [@lee1980;@alb2024] is used to connect points so that edges indicate neighbouring observations, in both the NLDR layout (@fig-NLDR-scurve (d)) and the \pD{} model representation. When the data has been binned the triangulation connectd centroids. The edges preserve the neighborhood information when the model is lifted into \pD{}.

<!-- When $k = 2$ Delaunay triangulation on $C^{(2)}$ generates the model in \gD{} space, which is a triangular mesh (@fig-NLDR-scurve (d)). It generates convex hulls of $C^{(2)}$ such that the circumcircle of every triangle in the triangulation contains no other points from $C^{(2)}$. -->

@@

<!-- langevitour with pD model for S-curve -->
::: {#fig-scurve-sc layout-ncol="4" fig-pos="H"}
![](figures/scurve/umap_trimesh_layout.png)

![](figures/scurve/sc_umap_best_1.png)

![](figures/scurve/sc_umap_best_2.png)

![](figures/scurve/sc_umap_best_3.png)

Model in \gD{}, on the UMAP layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data ($(s_1, \ s_2) = (-0.160, \ -0.263)$, $b = 405 \  (15, \ 27)$, $m = 70$, benchmark value to remove low density hexagons is $0.250$, and benchmark value to remove large edges is $0.159$). MSE is $0.0462$. The model closely fits the shape, but it doesn't fully fill out the width of the S, which means that it does not adequately capture the surface (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/i7F7xpN1Hz8>).
:::

@@

```{r}
#| label: best-umap-model-scurve

# Initialize effective bins along x
effective_bin1_scurve <- 15

# Fit high-dimensional model
scurve_model <- fit_highd_model(
  training_data = training_data_scurve,
  emb_df = umap_scurve_scaled,
  bin1 = effective_bin1_scurve,
  r2 = r2,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

df_bin_centroids_scurve <- scurve_model$df_bin_centroids
df_bin_scurve <- scurve_model$df_bin

hex_grid_nonempty <- hex_grid |>
  filter(hex_poly_id %in% df_bin_centroids_scurve$hexID)

```{r}
#| label: p-d-error-in-2d-scurve

## Compute error
error_df_scurve <- augment(
  df_bin_centroids = df_bin_centroids_scurve,
  df_bin = df_bin_scurve,
  training_data = training_data_scurve,
  newdata = NULL,
  type_NLDR = "UMAP",
  col_start = "x")

# error_df_scurve <- error_df_scurve |>
#   mutate(type = case_when(
#     row_wise_abs_error == 0 ~ "no error",
#     row_wise_abs_error <= 0.2 ~ "error 0-0.2",
#     row_wise_abs_error <= 0.4 ~ "error 0.2-0.4",
#     row_wise_abs_error <= 0.6 ~ "error 0.4-0.6",
#     .default = "error greter than 0.6"
#   )) |>
#   mutate(type = factor(type, levels = c(
#     "no error", "error 0-0.2", "error 0.2-0.4", "error 0.4-0.6", "error greter than 0.6")))

error_df_scurve <- error_df_scurve |>
  mutate(sqrt_row_wise_abs_error = sqrt(row_wise_abs_error))

error_df_scurve <- error_df_scurve |>
  bind_cols(umap_scurve_scaled |>
              select(-ID))

error_plot_scurve <- error_df_scurve |>
  ggplot(aes(x = UMAP1,
             y = UMAP2,
             colour = sqrt_row_wise_abs_error)) +
  geom_point(alpha=0.5) +
  scale_colour_continuous_sequential(palette = "Light Grays") +
  coord_equal()
```

```{r}
#| label: fig-p-d-error-in-2d-scurve
#| fig-cap: "error"
#| out-height: 30%
#| fig-pos: H

error_plot_scurve
```

@@

<!--prediction for true-model of S-curve-->
```{r}
#| label: model-prediction-umap-original

predict_umap_scurve <- read_rds(file = "data/s_curve/s_curve_umap_predict_true.rds")

predict_scurve_obj <- gen_scaled_data(
    data = predict_umap_scurve)

predict_umap_scurve_scaled <- predict_scurve_obj$scaled_nldr

plot_predict_umap <- ggplot(data = predict_umap_scurve_scaled,
                            aes(
                              x = UMAP1,
                              y = UMAP2
                            )) +
  geom_point(alpha = 0.5) +
  interior_annotation("b", c(0.92, 0.96))
```

```{r}
#| label: model-prediction-umap-quollr

true_pred_df <- predict_emb(
  test_data = true_model_df,
  df_bin_centroids = df_bin_centroids_scurve,
  df_bin = df_bin_scurve,
  type_NLDR = "UMAP"
)

# Compute radius r
a1 <- calc_bins_y(bin1 = 15, r2 = r2, q = 0.16)$a1
r <- a1 / 2

# Function to jitter points within a circumcircle of a hexagon
jitter_within_circumcircle <- function(center_x, center_y, radius, num_points) {
  theta <- runif(num_points, 0, 2 * pi)
  r <- radius * sqrt(runif(num_points))
  x <- center_x + r * cos(theta)
  y <- center_y + r * sin(theta)
  jittered_points <- data.frame(UMAP1 = x, UMAP2 = y, ID = points_in_hex$ID)
  return(jittered_points)
}

# Jittered points data frame
jittered_points_df <- data.frame()

# Iterate through each hexagon and jitter points within the circumcircle
for (hex_id in unique(true_pred_df$pred_hb_id)) {
  # hex_points <- hex_grid_nonempty %>% filter(hex_poly_id == hex_id)
  # center_x <- mean(hex_points$x)
  # center_y <- mean(hex_points$y)
  # points_in_hex <- true_pred_df %>% filter(pred_UMAP_1 >= min(hex_points$x) & pred_UMAP_1 <= max(hex_points$x) & pred_UMAP_2 >= min(hex_points$y) & pred_UMAP_2 <= max(hex_points$y))

  points_in_hex <- true_pred_df |>
    filter(pred_hb_id == hex_id)

  center_x <- points_in_hex |> 
    pull(pred_UMAP_1) |> 
    unique()

  center_y <- points_in_hex |> 
    pull(pred_UMAP_2) |> 
    unique()

  if (nrow(points_in_hex) > 0) {
    jittered_points <- jitter_within_circumcircle(center_x, center_y, r, nrow(points_in_hex))
    jittered_points_df <- rbind(jittered_points_df, jittered_points)
  }
}

plot_predict_umap_model <- ggplot(data = jittered_points_df,
                            aes(
                              x = UMAP1,
                              y = UMAP2
                            )) +
  geom_point(alpha = 0.5) +
  interior_annotation("c", c(0.92, 0.96))
```

```{r}
#| label: true-model-projection-langevitour

true_model_df_rm_id <- true_model_df |>
  select(-ID)

projection <- cbind(
    c(0.18390,-0.18443,-0.06860,0.00654,-0.02921,0.02313,-0.08963),
    c(-0.02902,0.03638,-0.16872,0.22418,0.01146,0.02777,0.01453))
projected_df <- as.matrix(true_model_df_rm_id) %*% projection |> 
  as_tibble(.name_repair = "unique") |> 
  rename(c("proj1" = "...1", 
           "proj2" = "...2"))

proj_plot_scurve_true <- ggplot(
  data = projected_df, 
  aes(
    x = proj1, 
    y = proj2)) +
    geom_point(alpha = 0.5) +
  interior_annotation("a", c(0.92, 0.96))  
```

```{r}
#| label: model-prediction-points-line

predict_umap_scurve_scaled <- predict_umap_scurve_scaled |>
  mutate(type = "predict_from_umap")

jittered_points_df <- jittered_points_df |>
  mutate(type = "predict_from_model")

predict_point_df <- bind_rows(
  predict_umap_scurve_scaled,
  jittered_points_df
)

plot_predict_umap_model_connect <- ggplot(data = predict_point_df,
                            aes(
                              x = UMAP1,
                              y = UMAP2, 
                              group = ID
                            )) +
  geom_point(alpha = 0.5) +
  geom_line(alpha = 0.3) +
  interior_annotation("d", c(0.92, 0.96))
```

```{r}
#| echo: false
#| label: fig-predict-scurve
#| fig-pos: H
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Comparison of prediction generated using the exiting `umap` prediction method and our method: (a) A view of the true model in projections from $7\\text{-}D$, (b) predicted data from the `umap` prediction method, (c) predicted data from our method, and (d) comparison of predictions from both methods. In plot (d), points representing the same data but predicted by different methods are connected by lines. Need to add concluded sentence after changing the data."

proj_plot_scurve_true + plot_predict_umap + 
  plot_predict_umap_model + plot_predict_umap_model_connect +
  plot_layout(ncol = 4)
```

<!-- ::: {#fig-scurve-pred-sc layout-ncol="2" fig-pos="H"} -->
<!-- ![](figures/scurve/sc_true_only.png){width="150" fig-align="center"} -->

<!-- ![](figures/scurve/pred_true_view.png){width="150" fig-align="center"} -->

@@

The values $(s_1, \ s_2)$ define the position of the centroid of the bottom left hexagon. By default, this is at $s_1 = -q, s_2 = -qr_2$, where $q$ is the buffer sound the data. The choice of these values can have some effect on the distribution of bin counts. @fig-param-scurve (a) illustrates this. The distribution of bin counts for $s_1$ varying between $-0.1-0.0$ is shown. Generally, a more uniform distribution among these possibilities would indicate that the bins are reliably capturing the underlying distribution of observations.

@@

<!--add three choice of bins 7, 11, 14-->

```{r}
#| label: hexbin-regular-scurve1
## hexagon binning to have regular hexagons
hb_obj_scurve1 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 7, 
  r2 = r2)

a1_1 <- calc_bins_y(
  bin1 = 7, 
  r2 = r2
)$a1

## Data set with all centroids
all_centroids_df1 <- hb_obj_scurve1$centroids

## Generate all coordinates of hexagons
hex_grid1 <- hb_obj_scurve1$hex_poly

## To obtain the standardise counts within hexbins
counts_df1 <- hb_obj_scurve1$std_cts
df_bin_centroids1 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df1, 
  counts_df = counts_df1) |>
  filter(drop_empty == FALSE) |>
  mutate(b1 = "b1 = 7")

hex_grid_with_counts_s_curve1 <- full_join(
  hex_grid1, 
  df_bin_centroids1 |> select(hexID, std_counts), 
    by = c("hex_poly_id" = "hexID"))

hex_grid_coloured_scurve1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve1, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
             aes(x = UMAP1, y = UMAP2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("a", position = c(0.92, 0.93)) +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.35, 2))

hex_grid_coloured_scurve1_dens <- ggplot(hex_grid_with_counts_s_curve1) +
  geom_histogram(aes(x=std_counts), 
                 breaks=seq(0, 1, 0.05),
                 fill="slategray4", colour="white")

>>>>

With our goal being to make a representation of this \gD{} layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first bin the data, reducing it from $n$ to $m\leq n$ observations, and connect the bin centroids. We recommend using a hexagon grid because it better reflects the data distribution and has less artifacts than a rectangular grid. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-two-curvy, and documented below.

@@

umap_two_curvy_scaled <- two_curvy_scaled_obj$scaled_nldr |>
  mutate(ID = row_number())
lim1 <- two_curvy_scaled_obj$lim1
lim2 <- two_curvy_scaled_obj$lim2
r2 <- diff(lim2)/diff(lim1)

sc_ltr_pos <- c(0.08, 0.93)
# sc_xlims <- c(-0.5, 1.47)
# sc_ylims <- c(-0.32, 2.1)
sc_xlims <- c(-0.2, 1.2)
sc_ylims <- c(-0.2, 0.8)

nldr_two_curvy <- umap_two_curvy_scaled |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos)
```

@@

```{r}
#| label: hexbin-two-curvy
## Compute hexbin parameters
num_bins_x_two_curvy <- 10

## hexagon binning to have regular hexagons
hb_obj_two_curvy <- hex_binning(
  data = umap_two_curvy_scaled, 
  bin1 = num_bins_x_two_curvy, 
  r2 = r2,
  q = 0.1)

## Data set with all centroids
all_centroids_df <- hb_obj_two_curvy$centroids

## Generate all coordinates of hexagons
hex_grid <- hb_obj_two_curvy$hex_poly

## To obtain the standardise counts within hexbins
counts_df <- hb_obj_two_curvy$std_cts
df_bin_centroids <- extract_hexbin_centroids(
  centroids_df = all_centroids_df, 
  counts_df = counts_df) |>
  filter(drop_empty == FALSE)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "black", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = umap_two_curvy_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5, size = 0.5, color = "#000000") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos)
```

@@

```{r}
#| label: empty-bin-two-curvy
hex_grid_nonempty <- hex_grid |>
  filter(hex_poly_id %in% df_bin_centroids$hexID)

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(color = "black", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids, 
             aes(x = c_x, y = c_y), 
             color = "#33a02c",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos) 
```

<!--2D model-->
```{r}
#| label: triangulate-two-curvy
## Triangulate bin centroids
tr1_object_two_curvy <- tri_bin_centroids(
  df_bin_centroids, x = "c_x", y = "c_y")
tr_from_to_df_two_curvy <- gen_edges(
  tri_object = tr1_object_two_curvy)

## Compute 2D distances
distance_two_curvy <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_two_curvy, 
  start_x = "x_from", 
  start_y = "y_from", 
  end_x = "x_to", 
  end_y = "y_to", 
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_two_curvy <- find_lg_benchmark(
  distance_edges = distance_two_curvy, 
  distance_col = "distance")

trimesh_removed_two_curvy <- vis_rmlg_mesh(
  distance_edges = distance_two_curvy, 
  benchmark_value = benchmark_two_curvy, 
  tr_coord_df = tr_from_to_df_two_curvy, 
  distance_col = "distance") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("d", sc_ltr_pos) 
```

@@

```{r}
#| label: fig-NLDR-two-curvy
#| echo: false
#| fig-cap: "Key steps for constructing the model on the UMAP layout ($k=2$): (a) data, (b) hexagon bins, (c) bin centroids, and (d) triangulated centroids. The S-curve data is shown."
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%

nldr_two_curvy + hex_grid_two_curvy + 
  hex_grid_nonempty_two_curvy + 
  trimesh_removed_two_curvy +
  plot_layout(ncol = 4)
```

<!--langevitour with true model for S-curve--> 
<!-- ::: {#fig-two_curvy-true-sc layout-ncol="3" fig-pos="H"} -->

<!-- ![](figures/two_curvy/sc_true_1.png){width="150" fig-align="center"} -->

<!-- ![](figures/two_curvy/sc_true_2.png){width="150" fig-align="center"} -->

<!-- ![](figures/two_curvy/sc_true_3.png){width="150" fig-align="center"} -->

@@

@@

Because we are working with distances between points, starting with data having a standard scale, e.g. [0, 1], is recommended. The default should take the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$ into account. When $k=2$, as in hexagon binning, the default range is $[0, y_{i,\text{max}}], i=1,2$, where $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = \frac{r_2}{r_1}$ (@fig-NLDR-two-curvy). If the NLDR aspect ratio is ignored then set $y_ {2,\text{max}} = 1$. <!-- \times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height ($a_1$) and width ($a_2$) of a regular hexagon.) The scaling of data should take the size of the hexagons into account, but choice of number of bins should. -->

@@

      y = -0.2, 

@@

      yend = -0.2), #-0.35),

@@

    annotate("text", x=0.5, y=-0.25, 

@@

      x = -0.2, 

@@

      xend = -0.2, 

@@

    annotate("text", x=-0.25, y=0.25, 

@@

    annotate("text", x=-0.18, y=-0.2, 

@@

When $k=2$, the number of bins along the $x$ and $y$ axes, $b_1$ and $b_2$, is computed by considering the scaling factors ${r_1, r_2}$, along with the offset along the axes ($q_1$, $q_2$), and the height ($a_1$) and width ($a_2$) of a regular hexagon (@fig-NLDR-two-curvy (b)). ($b_1 = \frac{r_1 + q_1}{a_2}$, and $b_2 = \frac{r_2 + q_2}{0.75 \times a_1}$) where $0.75 \times a_1$ accommodate to have hexagons fill the entire 2D space without leaving any gaps between them.)-->

@@

By default, the bin centroid is used for describing a hexagon (as done in @fig-NLDR-two-curvy (c)), but any measure of center, such as a mean or weighted mean of the points within each hexagon, could be used. The bin centers, and the binned data, are the two important components needed to render the model representation in high dimensions.

@@

Delaunay triangulation [@lee1980;@alb2024] is used to connect points so that edges indicate neighbouring observations, in both the NLDR layout (@fig-NLDR-two-curvy (d)) and the \pD{} model representation. When the data has been binned the triangulation connectd centroids. The edges preserve the neighborhood information when the model is lifted into \pD{}.

<!-- When $k = 2$ Delaunay triangulation on $C^{(2)}$ generates the model in \gD{} space, which is a triangular mesh (@fig-NLDR-two-curvy (d)). It generates convex hulls of $C^{(2)}$ such that the circumcircle of every triangle in the triangulation contains no other points from $C^{(2)}$. -->

@@

@@

@@

<!-- ::: {#fig-two_curvy-pred-sc layout-ncol="2" fig-pos="H"} -->
<!-- ![](figures/two_curvy/sc_true_only.png){width="150" fig-align="center"} -->

<!-- ![](figures/two_curvy/pred_true_view.png){width="150" fig-align="center"} -->

@@

The values $(s_1, \ s_2)$ define the position of the centroid of the bottom left hexagon. By default, this is at $s_1 = -q, s_2 = -qr_2$, where $q$ is the buffer sound the data. The choice of these values can have some effect on the distribution of bin counts. @fig-param-two_curvy (a) illustrates this. The distribution of bin counts for $s_1$ varying between $-0.1-0.0$ is shown. Generally, a more uniform distribution among these possibilities would indicate that the bins are reliably capturing the underlying distribution of observations.

@@

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \sqrt{\frac{n}{r_2}}$ are allowed. @fig-param-two_curvy (b) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

<!-- To determine the effective $b$, candidate values are selected based on the range between the minimum and approximate maximum $b_1$, because $b_2$ is computed from $b_1$. The minimum $b_1$ is set to $2$, while the maximum number is estimated by taking the square root of $\frac{n}{2}$. By evaluating MSE across varying $b$ within this range for different $q$, helps to determine an appropriate values for $b$ and $q$ (@fig-param-two_curvy (a)).-->
